Analysis of finite-volume advection schemes on cubed-sphere grids and an accurate scheme for divergent winds

### 2024-05-24 paper.tex
authorship contribution statement
removed highlights in red

\title{Analysis of finite-volume advection schemes on cubed-sphere grids and an accurate \textcolor{red}{scheme} for divergent winds}

@@

%% Keywords
\begin{keyword}
Cubed-sphere,
finite-volume,
transport,
advection,
\textcolor{red}{
atmosphere dynamics,
numerical weather prediction,
divergent winds}.
\end{keyword}

@@

\textcolor{red}{
The transport scheme employed in FV3, proposed by \cite{putman:2007}, extends the method introduced by \cite{lin:1996}, moving from latitude-longitude grids to cubed-sphere grids, aiming for better performance on massive parallel supercomputers, which is difficult to achieve on latitude-longitude grids due to the pole problem \cite{will:2007,stan:2012}.
Cubed-sphere grids were originally proposed by \cite{sadourny:1972} and revisited by \cite{rancic:1996,ronchi:1996}. 
This type of grid is an instance of grids based on Platonic solids \cite{stan:2012}, which consider a Platonic solid circumscribed on the sphere, project its faces onto the sphere's surface, and apply subdivision on the projected faces to generate the grid cells.
}

@@

\textcolor{red}{
Therefore, given the relevance of the FV3 dynamical core,
the goal of this study is to reassess and suggest enhancements for the current FV3 transport scheme originally developed by \cite{putman:2007}, given its pivotal role in the horizontal dynamics of FV3.}
It is demonstrated in this work that the scheme proposed by \cite{putman:2007} assumes constant metric terms during the application of the PPM to each 1D flux integration domain and employs a first-order departure point calculation for the 1D fluxes.
\textcolor{red}{
A scheme that incorporates a second-order departure point calculation for the 1D fluxes and eliminates the assumption of constant metric terms is proposed.}
While the proposed scheme does not retain the property of splitting error elimination for divergence-free winds and constant scalar fields, it ensures second-order errors in such scenarios.

@@

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Cubed-sphere grids}
\label{cs-grids}
The goal of this Section is to briefly review the concept of cubed-sphere grids, particularly those available in FV3.
To start with, the mapping between the cube and the sphere introduced by \cite{sadourny:1972}, also known as the equidistant mapping, is presented in Section \ref{cs-equidistant}.
After that, it is shown that by using a change of coordinates, the equidistant mapping can be utilized to create other mappings from the cube to the sphere.
Namely, Section \ref{cs-equiangular} introduces the equiangular mapping introduced by \cite{ronchi:1996}, and Section \ref{cs-equiedge} introduces the equi-edge mapping introduced by \cite{chen:2021}. 
Section \ref{cs-grid} demonstrates how these mappings are utilized to generate cubed-sphere grids and introduces all the notations and it introduces all the notations and tools necessary for the subsequent parts of this work.
\textcolor{red}{
Finally, on the cubed-sphere, it is needed to define ghost cells, which are cells added outside each cubed-sphere face that allow for stencil computation near to the edges of the spherical cube.
Therefore, in Section \ref{cs-ghost}, it is demonstrated how the needed ghost cells are generated.
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Equidistant mapping}
\label{cs-equidistant}
Considering a cube circumscribed on a sphere, \cite{sadourny:1972} introduces a mapping between the cube faces and the sphere, generating a spherical cube.
This mapping is also called equidistant mapping and generates the equidistant grid.
Given $R>0$, the sphere of radius $R$ 
centered at the origin of  $\mathbb{R}^3$ is denoted as:
\begin{equation}
	\label{s2_r}
	\mathbb{S}^2_R = \{ P = (p_x,p_y,p_z) \in \mathbb{R}^3: p_x^2 + p_y^2 + p_z^2 = R^2\}.
\end{equation}
For the purposes of this work, the Earth radius $R=6.371\times 10^6$ meters is considered.
The equidistant mapping considers a cube centered at the origin with a side length of $\frac{2R}{\sqrt{3}}$ and radially projects the cube faces onto the sphere (Figure \ref{sph-cube-equidist}).

The equidistant mapping is a family of maps ${\Gamma}_{p}: [-1,1] \times [-1,1] \to \mathbb{S}^2_R$, where $p=1, \ldots, 6$, defined as follows:
\begin{align}
	\label{gamma_1}
	{\Gamma}_{1}(X,Y) &= \frac{R}{\sqrt{1 + X^2 + Y^2}}(1, X, Y),\\ 
	\label{gamma_2}
	{\Gamma}_{2}(X,Y) &= \frac{R}{\sqrt{1 + X^2 + Y^2}}(-X, 1, Y), \\
	\label{gamma_3}
	{\Gamma}_{3}(X,Y) &= \frac{R}{\sqrt{1 + X^2 + Y^2}}(-X, -Y, 1), \\
	\label{gamma_456}
{\Gamma}_{3+k}(X,Y)&=-{\Gamma}_{k}(Y,X),\quad k=1,2,3.
\end{align}
\textcolor{red}{
These mappings are defined individually for each of the 6 cube faces, also called panels, denoted by $p$, and they allow for the coverage of the sphere.}
The idea behind this mapping is illustrated in Figure \ref{sph-cube-equidist}, where the coordinates $(X,Y)$ are thought to live on the cube faces.
Figure \ref{sph-cube-equidist} shows how the grid points are equally spaced on the cube and then projected onto the sphere, hence the name equidistant.
Note that there are other ways to arrange the coordinates over the panels.
Each one defines a \textcolor{red}{connectivity pattern} between the panels, as discussed by \cite[Section 2.1]{chen:2021}.
The \textcolor{red}{connectivity pattern} presented here is known as the staircase arrangement \textcolor{red}{(see Figure 2 in \cite{chen:2021})}, which is used in FV3 and provides some advantages for exchanging information between panels.
\begin{figure}[!htb]
	\centering
	\begin{subfigure}{0.45\textwidth}
		\centering
		\includegraphics[width=1.1\linewidth]{g1}
		\caption{Cube and sphere equidistant mapping for $p_z=0$.\label{sph-cube-equidist}}
	\end{subfigure}

	\begin{subfigure}{0.45\textwidth}
		\centering
		\includegraphics[width=1.1\linewidth]{g2}
		\caption{Equiangular mapping for $p_z=0$.\label{sph-cube-equiangular}}
	\end{subfigure}
	\begin{subfigure}{0.45\textwidth}
	\centering
	\includegraphics[width=1.1\linewidth]{g0}
	\caption{\textcolor{red}{Equi-edge mapping for $p_z=0$.}\label{sph-cube-equiedge}}
\end{subfigure}
\caption{Illustration of the cube-to-sphere projection using the equidistant (a),  equiangular (b) and the \textcolor{red}{equi-edge}  (c) mappings. This figure uses a cross section obtained with $p_z=0$.}
\label{sph-cube-mappings}
\end{figure}

@@

\textcolor{red}{Given a tangent vector field $\boldsymbol{u}$ on the sphere, also known as wind, we may represent it using the basis obtained by cubed-sphere coordinates:
\begin{equation}
	\label{contravariant-wind}
	\boldsymbol{u}(x, y, p) = 
	\mathfrak{u}(x, y,p ) \partial_x{\Psi}_p(x, y) + 
	\mathfrak{v}(x, y, p) \partial_y{\Psi}_p(x, y).
\end{equation}
This representation $(\mathfrak{u},\mathfrak{v})$ is known as the contravariant representation, and they are the components of the wind on the tangent basis defined by the cubed-sphere mapping.
A detailed discussion on how the cubed-sphere wind representation is related to the zonal and meridional representation is presented in \ref{ap-wind}.
In practice, FV3 schemes \cite{putman:2007, harris:2021} use the normalized contravariant wind
$({u},{v})$ given by:
\begin{equation}
	\label{norm-contravariant-wind}
	\boldsymbol{u}(x, y, p) = 
	{u}(x, y, p) \boldsymbol{e}_x(x, y, p) + 
	{v}(x, y, p) \boldsymbol{e}_y(x, y, p),
\end{equation}
where $\boldsymbol{e}_x$ and $\boldsymbol{e}_y$ are the normalized cubed-sphere tangent vectors, which may be computed easily in terms of the grid points \cite[Appendix C2]{chen:2021}.
It is easy to see that:
\begin{equation}
	\label{contra-uv}
	\mathfrak{u}(x,y,p)  = \frac{{u}(x,y,p)}{\|\partial_x{\Psi}_p(x,y)\|}, \quad
	\mathfrak{v}(x,y,p)  = \frac{{v}(x,y,p)}{\|\partial_y{\Psi}_p(x,y)\|}.
\end{equation}
Finally, we recall that the horizontal divergence operator for a  wind $\boldsymbol{u}$  on the sphere is defined in terms of the cubed-sphere metric terms as follows:
\begin{align}
	\label{div-def}
	[\nabla \cdot {\boldsymbol{u}}](x,y,p) :=
	\frac{1}{\sqrt{\mathfrak{g}_{\Psi}}(x,y)}
	\bigg(
	\partial_x(\sqrt{\mathfrak{g}_{\Psi}}\mathfrak{u})(x,y,p) +
	\partial_y(\sqrt{\mathfrak{g}_{\Psi}}\mathfrak{v})(x,y,p)
	\bigg),
\end{align}
for $x,y\in[-a,a]$, $p$ is the panel and
$\mathfrak{u}$ and $\mathfrak{v}$ are the contravariant wind components.
The divergence operator shall be used in the transport model in Section \ref{adv-2d}.
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Equiangular mapping}
\label{cs-equiangular}
Another cubed-sphere mapping is the equiangular mapping, 
introduced by \cite{ronchi:1996}, which leads to a more uniform grid on the sphere.
This mapping is obtained by considering $\beta(x) = \tan{x}$ and $a=\frac{\pi}{4}$.
In this case, $\beta(x)$ represents the angular coordinates, and the cube-sphere is obtained by partitioning the angle between
grid points equally, as illustrated in Figure \ref{sph-cube-equiangular}, hence the name equiangular.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Equi-edge mapping}
\label{cs-equiedge}
\textcolor{red}{
Another cubed-sphere mapping is the equi-edge mapping, initially introduced by \cite{chen:2021}, which utilizes $\beta(x) = \sqrt{2}\tan{x}$ and $a=\arcsin{\big(\frac{1}{\sqrt{3}}\big)}$. It is worth noting that while this mapping technique had been used previously in FV3, it was not formally documented until the work \cite{chen:2021}.}
The idea behind the equi-edge mapping lies in partitioning the edges of the spherical cube equally,
and then generating the other cells, hence the name equi-edge.
Also, this mapping leads to more uniform grid cells after applying the grid stretching option of FV3 \citep{harris:2016, chen:2021}.
\textcolor{red}{This mapping is illustrated in Figure \ref{sph-cube-equiedge}}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Cubed-sphere grid generation}
\label{cs-grid}
Let us fix two positive integers $N$ and $\nu$, where $N$ represents the number of cells in both the $x$ and $y$ directions, and $\nu$ represents the number of ghost cell layers. 
The equiangular or equi-edge mappings, denoted by $\Psi_p$, introduced previously, are considered to generate the cubed-sphere grid.
\textcolor{red}{The notation $\Psi_p$ is used for both equiangular and equi-edge mappings, as what will be discussed does not depend particularly on the mapping.}
For simplicity, the metric term $\sqrt{\mathfrak{g}}_{\Psi}$ is denoted by $\sqrt{\mathfrak{g}}$.
To generate the cubed-sphere grid, the domain $[-a,a]\times[-a,a]$ is discretized using uniformly spaced points.
\begin{align}
\label{edge-coords}
x_{i+\frac{1}{2}} = -a+i\Delta x, \quad
y_{j+\frac{1}{2}} = -a+j\Delta y,
\end{align}
where $\Delta x = \Delta y = \frac{2a}{N}$, $i,j=-\nu+1,\ldots,N+1+\nu$. The center coordinates  are defined as:
\begin{align}
	\label{center-coords}
	x_{i} = \frac{x_{i+\frac{1}{2}}+x_{i-\frac{1}{2}}}{2}, \quad
	y_{j} = \frac{y_{j+\frac{1}{2}}+y_{j-\frac{1}{2}}}{2},
\end{align}
for $i,j=-\nu+1,\ldots,N+\nu$.
Notice that the mappings ${\Psi}_p$ defined before can be computed outside the range $[-a,a]$, and the cubed-sphere mapping can be applied to all these ghost cell points.
Firstly, we shall focus the attention on the interior cells; the generation of ghost cells shall be addressed in Section \ref{cs-ghost}.

\begin{figure}[!htb]
	\centering
	\includegraphics[width=0.5\linewidth]{fv_grid}
	\caption{
	\textcolor{red}{
	Illustration of the discrete grid indexes for a cell. 
	The corner points are depicted using black squares, while the centroids are represented using black circles. 
	The  left-right and up-down midpoints of the edges,  are illustrated in blue and red circles, respectively.
Additionally, this figure shows the positions of the contravariant wind components $\mathfrak{u}$ and $\mathfrak{v}$ in a C-grid discretization for the transport model, along with the fluid density $\rho$ and tracer concentration $\phi$ at the centers.}
	\label{cgrid}}
\end{figure}

There are four types of grid points on the cubed-sphere that are needed to be computed: the center, corners, right-left edge midpoints, and  up-down edge midpoints. 
\textcolor{red}
{These points are illustrated in Figure \ref{cgrid}.}
The corner points are computed as:
\begin{equation}
	\label{corner-points}
	\Psi_{i+\frac{1}{2},j+\frac{1}{2},p} := {\Psi}_p(x_{i+\frac{1}{2}},y_{j+\frac{1}{2}}),
\end{equation}
$i,j=0, \ldots, N$.
To ease the notation hereafter, the dependence on $p$ is omitted because it does not interfere with what is going to be discussed in this Section.
Figure \ref{cs-grids-N10} shows the obtained grid lines in for $N=10$.
\begin{figure}[!htb]
	\begin{subfigure}{0.42\textwidth}
		\centering
		\includegraphics[width=1.1\linewidth]{gnomonic_equiedge_cs_10_sphere}
		\caption{Equiangular grid}
	\end{subfigure}
	\centering
\begin{subfigure}{0.42\textwidth}
	\centering
	\includegraphics[width=1.1\linewidth]{gnomonic_equiangular_cs_10_sphere}
	\caption{Equi-edge grid}
\end{subfigure}
	\caption{(a) Illustration of the resulting gridlines
		for the cubed-sphere equiangular and equi-edge mapping for $N=10$.\label{cs-grids-N10}}
\end{figure}

@@

\section{The conservative transport equation on the cubed-sphere}
\label{adv-2d}
The goal of this Section is to present and solve numerically the conservative transport equation on the cubed-sphere. 
We are going to consider the equi-edge or equiangular cubed-sphere mappings $\Psi_p$ and their respective local coordinates $(x,y)$.
\textcolor{red}{Once again, the metric term is denoted by $\sqrt{\mathfrak{g}}_{\Psi}$ by $\sqrt{\mathfrak{g}}$ for simplicity.}
Following \cite{nair:2010}, the transport model without sources or sinks is considered:
\begin{align}
	\label{transp-rho}
	[\partial_t \rho + \nabla \cdot (\rho \boldsymbol{u})](x,y,p,t)&=0,\\
	\label{transp-phi}
	[\partial_t (\rho \phi) + \nabla \cdot (\rho \phi \boldsymbol{u})](x,y,p,t)&=0,
\end{align}
for $x,y\in[-a,a]$, $t \in [0,T]$, where $T$ is the final time,
\textcolor{red}
{$\boldsymbol{u}$ is the wind, $\mathfrak{u}$ and $\mathfrak{v}$ are the contravariant wind components (Equation \ref{contravariant-wind}),
$\rho$ is the fluid density and $\phi$ is the tracer concentration}.
Additionally, for the transport model, it is assumed that the initial conditions are given as $\rho(x,y,p,0)=\rho_0(x,y,p)$ and $\phi(x,y,p,0)=\phi_0(x,y,p)$, given $\rho_0$ and $\phi_0$. % and that the boundary conditions are related to the cubed-sphere topology.
Equation \eqref{transp-rho} is the continuity equation and Equation \eqref{transp-phi} is the conservative advection equation.
In this framework, $\rho \phi$ represents the tracer density, and the total masses of $\rho$ and $\rho \phi$ are preserved.  Therefore, these quantities are referred to as conserved quantities.

In the transport model, one can easily see that the tracer variable $\phi$ is advected.
That is, $\phi$ satisfies the non-conservative advection equation:
\begin{align}
	\label{adv-phi}
	[\partial_t \phi +  
	\langle\boldsymbol{u}, \nabla \phi \rangle](x,y,p,t)&=0.
\end{align}
Equations \eqref{transp-rho} and \eqref{transp-phi} have the same form. 
\textcolor{red}{Hence, it follows from the definition of the divergence operator in terms of the cubed-sphere mapping (Equation \eqref{div-def}) that the equation that needs to be solved may be uniquely expressed as:}
\begin{equation}
	\label{2d-adv}
	[{\partial_t (\sqrt{\mathfrak{g}}q)}+
	{\partial_x (\mathfrak{u}\sqrt{\mathfrak{g}}q)}+
	{\partial_y (\mathfrak{v}\sqrt{\mathfrak{g}}q)}]
	(x, y, p, t) = 0,
\end{equation}
where $q=\rho$ or $q=\rho \phi$. 
Additionally, it is assumed that $q(x,y,p,0)=q_0(x,y,p)$ for some given $q_0$.
Equation \eqref{2d-adv} is referred to as the conservative transport equation.
The goal now is to solve Equation \eqref{2d-adv}, which will allow for the solution of the transport model on the sphere.
In the shallow-water model, Equation \eqref{2d-adv} is satisfied for the fluid depth and the absolute vorticity.

>>>>

\title{Analysis of finite-volume advection schemes on cubed-sphere grids and an accurate scheme for divergent winds}

@@

%% Keywords
\begin{keyword}
Cubed-sphere,
finite-volume,
transport,
advection,
atmosphere dynamics,
numerical weather prediction,
divergent winds.
\end{keyword}

@@

The transport scheme employed in FV3, proposed by \cite{putman:2007}, extends the method introduced by \cite{lin:1996}, moving from latitude-longitude grids to cubed-sphere grids, aiming for better performance on massive parallel supercomputers, which is difficult to achieve on latitude-longitude grids due to the pole problem \cite{will:2007,stan:2012}.
Cubed-sphere grids were originally proposed by \cite{sadourny:1972} and revisited by \cite{rancic:1996,ronchi:1996}. 
This type of grid is an instance of grids based on Platonic solids \cite{stan:2012}, which consider a Platonic solid circumscribed on the sphere, project its faces onto the sphere's surface, and apply subdivision on the projected faces to generate the grid cells.

@@

Therefore, given the relevance of the FV3 dynamical core,
the goal of this study is to reassess and suggest enhancements for the current FV3 transport scheme originally developed by \cite{putman:2007}, given its pivotal role in the horizontal dynamics of FV3.
It is demonstrated in this work that the scheme proposed by \cite{putman:2007} assumes constant metric terms during the application of the PPM to each 1D flux integration domain and employs a first-order departure point calculation for the 1D fluxes.
A scheme that incorporates a second-order departure point calculation for the 1D fluxes and eliminates the assumption of constant metric terms is proposed.
While the proposed scheme does not retain the property of splitting error elimination for divergence-free winds and constant scalar fields, it ensures second-order errors in such scenarios.

@@

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Cubed-sphere grids}
\label{cs-grids}
The goal of this Section is to briefly review the concept of cubed-sphere grids, particularly those available in FV3.
To start with, the mapping between the cube and the sphere introduced by \cite{sadourny:1972}, also known as the equidistant mapping, is presented in Section \ref{cs-equidistant}.
After that, it is shown that by using a change of coordinates, the equidistant mapping can be utilized to create other mappings from the cube to the sphere.
Namely, Section \ref{cs-equiangular} introduces the equiangular mapping introduced by \cite{ronchi:1996}, and Section \ref{cs-equiedge} introduces the equi-edge mapping introduced by \cite{chen:2021}. 
Section \ref{cs-grid} demonstrates how these mappings are utilized to generate cubed-sphere grids and introduces all the notations and it introduces all the notations and tools necessary for the subsequent parts of this work.
Finally, on the cubed-sphere, it is needed to define ghost cells, which are cells added outside each cubed-sphere face that allow for stencil computation near to the edges of the spherical cube.
Therefore, in Section \ref{cs-ghost}, it is demonstrated how the needed ghost cells are generated.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Equidistant mapping}
\label{cs-equidistant}
Considering a cube circumscribed on a sphere, \cite{sadourny:1972} introduces a mapping between the cube faces and the sphere, generating a spherical cube.
This mapping is also called equidistant mapping and generates the equidistant grid.
Given $R>0$, the sphere of radius $R$ 
centered at the origin of  $\mathbb{R}^3$ is denoted as:
\begin{equation}
	\label{s2_r}
	\mathbb{S}^2_R = \{ P = (p_x,p_y,p_z) \in \mathbb{R}^3: p_x^2 + p_y^2 + p_z^2 = R^2\}.
\end{equation}
For the purposes of this work, the Earth radius $R=6.371\times 10^6$ meters is considered.
The equidistant mapping considers a cube centered at the origin with a side length of $\frac{2R}{\sqrt{3}}$ and radially projects the cube faces onto the sphere (Figure \ref{sph-cube-equidist}).

The equidistant mapping is a family of maps ${\Gamma}_{p}: [-1,1] \times [-1,1] \to \mathbb{S}^2_R$, where $p=1, \ldots, 6$, defined as follows:
\begin{align}
	\label{gamma_1}
	{\Gamma}_{1}(X,Y) &= \frac{R}{\sqrt{1 + X^2 + Y^2}}(1, X, Y),\\ 
	\label{gamma_2}
	{\Gamma}_{2}(X,Y) &= \frac{R}{\sqrt{1 + X^2 + Y^2}}(-X, 1, Y), \\
	\label{gamma_3}
	{\Gamma}_{3}(X,Y) &= \frac{R}{\sqrt{1 + X^2 + Y^2}}(-X, -Y, 1), \\
	\label{gamma_456}
{\Gamma}_{3+k}(X,Y)&=-{\Gamma}_{k}(Y,X),\quad k=1,2,3.
\end{align}

These mappings are defined individually for each of the 6 cube faces, also called panels, denoted by $p$, and they allow for the coverage of the sphere.
The idea behind this mapping is illustrated in Figure \ref{sph-cube-equidist}, where the coordinates $(X,Y)$ are thought to live on the cube faces.
Figure \ref{sph-cube-equidist} shows how the grid points are equally spaced on the cube and then projected onto the sphere, hence the name equidistant.
Note that there are other ways to arrange the coordinates over the panels.
Each one defines a connectivity pattern between the panels, as discussed by \cite[Section 2.1]{chen:2021}.
The connectivity pattern presented here is known as the staircase arrangement (see Figure 2 in \cite{chen:2021}), which is used in FV3 and provides some advantages for exchanging information between panels.
\begin{figure}[!htb]
	\centering
	\begin{subfigure}{0.45\textwidth}
		\centering
		\includegraphics[width=1.1\linewidth]{g1}
		\caption{Cube and sphere equidistant mapping for $p_z=0$.\label{sph-cube-equidist}}
	\end{subfigure}

	\begin{subfigure}{0.45\textwidth}
		\centering
		\includegraphics[width=1.1\linewidth]{g2}
		\caption{Equiangular mapping for $p_z=0$.\label{sph-cube-equiangular}}
	\end{subfigure}
	\begin{subfigure}{0.45\textwidth}
	\centering
	\includegraphics[width=1.1\linewidth]{g0}
	\caption{Equi-edge mapping for $p_z=0$.\label{sph-cube-equiedge}}
\end{subfigure}
\caption{Illustration of the cube-to-sphere projection using the equidistant (a),  equiangular (b) and the equi-edge  (c) mappings. This figure uses a cross section obtained with $p_z=0$.}
\label{sph-cube-mappings}
\end{figure}

@@

Given a tangent vector field $\boldsymbol{u}$ on the sphere, also known as wind, we may represent it using the basis obtained by cubed-sphere coordinates:
\begin{equation}
	\label{contravariant-wind}
	\boldsymbol{u}(x, y, p) = 
	\mathfrak{u}(x, y,p ) \partial_x{\Psi}_p(x, y) + 
	\mathfrak{v}(x, y, p) \partial_y{\Psi}_p(x, y).
\end{equation}
This representation $(\mathfrak{u},\mathfrak{v})$ is known as the contravariant representation, and they are the components of the wind on the tangent basis defined by the cubed-sphere mapping.
A detailed discussion on how the cubed-sphere wind representation is related to the zonal and meridional representation is presented in \ref{ap-wind}.
In practice, FV3 schemes \cite{putman:2007, harris:2021} use the normalized contravariant wind
$({u},{v})$ given by:
\begin{equation}
	\label{norm-contravariant-wind}
	\boldsymbol{u}(x, y, p) = 
	{u}(x, y, p) \boldsymbol{e}_x(x, y, p) + 
	{v}(x, y, p) \boldsymbol{e}_y(x, y, p),
\end{equation}
where $\boldsymbol{e}_x$ and $\boldsymbol{e}_y$ are the normalized cubed-sphere tangent vectors, which may be computed easily in terms of the grid points \cite[Appendix C2]{chen:2021}.
It is easy to see that:
\begin{equation}
	\label{contra-uv}
	\mathfrak{u}(x,y,p)  = \frac{{u}(x,y,p)}{\|\partial_x{\Psi}_p(x,y)\|}, \quad
	\mathfrak{v}(x,y,p)  = \frac{{v}(x,y,p)}{\|\partial_y{\Psi}_p(x,y)\|}.
\end{equation}
Finally, we recall that the horizontal divergence operator for a  wind $\boldsymbol{u}$  on the sphere is defined in terms of the cubed-sphere metric terms as follows:
\begin{align}
	\label{div-def}
	[\nabla \cdot {\boldsymbol{u}}](x,y,p) :=
	\frac{1}{\sqrt{\mathfrak{g}_{\Psi}}(x,y)}
	\bigg(
	\partial_x(\sqrt{\mathfrak{g}_{\Psi}}\mathfrak{u})(x,y,p) +
	\partial_y(\sqrt{\mathfrak{g}_{\Psi}}\mathfrak{v})(x,y,p)
	\bigg),
\end{align}
for $x,y\in[-a,a]$, $p$ is the panel and
$\mathfrak{u}$ and $\mathfrak{v}$ are the contravariant wind components.
The divergence operator shall be used in the transport model in Section \ref{adv-2d}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Equiangular mapping}
\label{cs-equiangular}
Another cubed-sphere mapping is the equiangular mapping, 
introduced by \cite{ronchi:1996}, which leads to a more uniform grid on the sphere.
This mapping is obtained by considering $\beta(x) = \tan{x}$ and $a=\frac{\pi}{4}$.
In this case, $\beta(x)$ represents the angular coordinates, and the cube-sphere is obtained by partitioning the angle between
grid points equally, as illustrated in Figure \ref{sph-cube-equiangular}, hence the name equiangular.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Equi-edge mapping}
\label{cs-equiedge}
Another cubed-sphere mapping is the equi-edge mapping, initially introduced by \cite{chen:2021}, which utilizes $\beta(x) = \sqrt{2}\tan{x}$ and $a=\arcsin{\big(\frac{1}{\sqrt{3}}\big)}$. It is worth noting that while this mapping technique had been used previously in FV3, it was not formally documented until the work \cite{chen:2021}.
The idea behind the equi-edge mapping lies in partitioning the edges of the spherical cube equally,
and then generating the other cells, hence the name equi-edge.
Also, this mapping leads to more uniform grid cells after applying the grid stretching option of FV3 \citep{harris:2016, chen:2021}.
This mapping is illustrated in Figure \ref{sph-cube-equiedge}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Cubed-sphere grid generation}
\label{cs-grid}
Let us fix two positive integers $N$ and $\nu$, where $N$ represents the number of cells in both the $x$ and $y$ directions, and $\nu$ represents the number of ghost cell layers. 
The equiangular or equi-edge mappings, denoted by $\Psi_p$, introduced previously, are considered to generate the cubed-sphere grid.
The notation $\Psi_p$ is used for both equiangular and equi-edge mappings, as what will be discussed does not depend particularly on the mapping.
For simplicity, the metric term $\sqrt{\mathfrak{g}}_{\Psi}$ is denoted by $\sqrt{\mathfrak{g}}$.
To generate the cubed-sphere grid, the domain $[-a,a]\times[-a,a]$ is discretized using uniformly spaced points.
\begin{align}
\label{edge-coords}
x_{i+\frac{1}{2}} = -a+i\Delta x, \quad
y_{j+\frac{1}{2}} = -a+j\Delta y,
\end{align}
where $\Delta x = \Delta y = \frac{2a}{N}$, $i,j=-\nu+1,\ldots,N+1+\nu$. The center coordinates  are defined as:
\begin{align}
	\label{center-coords}
	x_{i} = \frac{x_{i+\frac{1}{2}}+x_{i-\frac{1}{2}}}{2}, \quad
	y_{j} = \frac{y_{j+\frac{1}{2}}+y_{j-\frac{1}{2}}}{2},
\end{align}
for $i,j=-\nu+1,\ldots,N+\nu$.
Notice that the mappings ${\Psi}_p$ defined before can be computed outside the range $[-a,a]$, and the cubed-sphere mapping can be applied to all these ghost cell points.
Firstly, we shall focus the attention on the interior cells; the generation of ghost cells shall be addressed in Section \ref{cs-ghost}.

\begin{figure}[!htb]
	\centering
	\includegraphics[width=0.5\linewidth]{fv_grid}
	\caption{
	Illustration of the discrete grid indexes for a cell. 
	The corner points are depicted using black squares, while the centroids are represented using black circles. 
	The  left-right and up-down midpoints of the edges,  are illustrated in blue and red circles, respectively.
Additionally, this figure shows the positions of the contravariant wind components $\mathfrak{u}$ and $\mathfrak{v}$ in a C-grid discretization for the transport model, along with the fluid density $\rho$ and tracer concentration $\phi$ at the centers.
	\label{cgrid}}
\end{figure}

There are four types of grid points on the cubed-sphere that are needed to be computed: the center, corners, right-left edge midpoints, and  up-down edge midpoints. 
These points are illustrated in Figure \ref{cgrid}.
The corner points are computed as:
\begin{equation}
	\label{corner-points}
	\Psi_{i+\frac{1}{2},j+\frac{1}{2},p} := {\Psi}_p(x_{i+\frac{1}{2}},y_{j+\frac{1}{2}}),
\end{equation}
$i,j=0, \ldots, N$.
To ease the notation hereafter, the dependence on $p$ is omitted because it does not interfere with what is going to be discussed in this Section.
Figure \ref{cs-grids-N10} shows the obtained grid lines in for $N=10$.
\begin{figure}[!htb]
	\begin{subfigure}{0.42\textwidth}
		\centering
		\includegraphics[width=1.1\linewidth]{gnomonic_equiedge_cs_10_sphere}
		\caption{Equiangular grid}
	\end{subfigure}
	\centering
\begin{subfigure}{0.42\textwidth}
	\centering
	\includegraphics[width=1.1\linewidth]{gnomonic_equiangular_cs_10_sphere}
	\caption{Equi-edge grid}
\end{subfigure}
	\caption{(a) Illustration of the resulting gridlines
		for the cubed-sphere equiangular and equi-edge mapping for $N=10$.\label{cs-grids-N10}}
\end{figure}

@@

\section{The conservative transport equation on the cubed-sphere}
\label{adv-2d}
The goal of this Section is to present and solve numerically the conservative transport equation on the cubed-sphere. 
We are going to consider the equi-edge or equiangular cubed-sphere mappings $\Psi_p$ and their respective local coordinates $(x,y)$.
Once again, the metric term is denoted by $\sqrt{\mathfrak{g}}_{\Psi}$ by $\sqrt{\mathfrak{g}}$ for simplicity.
Following \cite{nair:2010}, the transport model without sources or sinks is considered:
\begin{align}
	\label{transp-rho}
	[\partial_t \rho + \nabla \cdot (\rho \boldsymbol{u})](x,y,p,t)&=0,\\
	\label{transp-phi}
	[\partial_t (\rho \phi) + \nabla \cdot (\rho \phi \boldsymbol{u})](x,y,p,t)&=0,
\end{align}
for $x,y\in[-a,a]$, $t \in [0,T]$, where $T$ is the final time,
$\boldsymbol{u}$ is the wind, $\mathfrak{u}$ and $\mathfrak{v}$ are the contravariant wind components (Equation \ref{contravariant-wind}),
$\rho$ is the fluid density and $\phi$ is the tracer concentration.
Additionally, for the transport model, it is assumed that the initial conditions are given as $\rho(x,y,p,0)=\rho_0(x,y,p)$ and $\phi(x,y,p,0)=\phi_0(x,y,p)$, given $\rho_0$ and $\phi_0$. % and that the boundary conditions are related to the cubed-sphere topology.
Equation \eqref{transp-rho} is the continuity equation and Equation \eqref{transp-phi} is the conservative advection equation.
In this framework, $\rho \phi$ represents the tracer density, and the total masses of $\rho$ and $\rho \phi$ are preserved.  Therefore, these quantities are referred to as conserved quantities.

In the transport model, one can easily see that the tracer variable $\phi$ is advected.
That is, $\phi$ satisfies the non-conservative advection equation:
\begin{align}
	\label{adv-phi}
	[\partial_t \phi +  
	\langle\boldsymbol{u}, \nabla \phi \rangle](x,y,p,t)&=0.
\end{align}
Equations \eqref{transp-rho} and \eqref{transp-phi} have the same form. 
Hence, it follows from the definition of the divergence operator in terms of the cubed-sphere mapping (Equation \eqref{div-def}) that the equation that needs to be solved may be uniquely expressed as:
\begin{equation}
	\label{2d-adv}
	[{\partial_t (\sqrt{\mathfrak{g}}q)}+
	{\partial_x (\mathfrak{u}\sqrt{\mathfrak{g}}q)}+
	{\partial_y (\mathfrak{v}\sqrt{\mathfrak{g}}q)}]
	(x, y, p, t) = 0,
\end{equation}
where $q=\rho$ or $q=\rho \phi$. 
Additionally, it is assumed that $q(x,y,p,0)=q_0(x,y,p)$ for some given $q_0$.
Equation \eqref{2d-adv} is referred to as the conservative transport equation.
The goal now is to solve Equation \eqref{2d-adv}, which will allow for the solution of the transport model on the sphere.
In the shallow-water model, Equation \eqref{2d-adv} is satisfied for the fluid depth and the absolute vorticity.
